### 2025-03-31 decision-making.qmd
20250330 - reinforcement schedules

>>>>

### Reinforcement Schedules {#sec-reinforcementSchedules}

Reinforcement schedules are depicted in @fig-reinforcementSchedules, as adapted from @Spielman2020.

::: {#fig-reinforcementSchedules}
![](images/reinforcementSchedules.png){width=100% fig-alt="Reinforcement Schedules. The four reinforcement schedules yield different response patterns. The variable ratio schedule is unpredictable and yields high and steady response rates, with little if any pause after reinforcement (e.g., gambler who plays the slot machines). A fixed ratio schedule is predictable and produces a high response rate, with a short pause after reinforcement (e.g., salesperson paid on commission). The variable interval schedule is unpredictable and produces a moderate, steady response rate (e.g., restaurant manager who is paid a bonus if the food inspector, who comes at unpredictable times, gives the restaurant a good rating). The fixed interval schedule yields a scallop-shaped response pattern, reflecting a significant pause after reinforcement (e.g., surgery patient who can receive a painkiller after pressing a button, but only up to once per hour). (Figure and caption adapted from Spielman, Lovett, & Jenkins (2020); https://openstax.org/books/psychology-2e/pages/6-3-operant-conditioning; Spielman, R. M., Jenkins, W. J., & Lovett, M. D. (2000). *Psychology*. OpenStax. https://openstax.org/details/books/psychology-2e; archived at https://perma.cc/9DZ8-34X2)"}

Reinforcement Schedules. The four reinforcement schedules yield different response patterns. The variable ratio schedule is unpredictable and yields high and steady response rates, with little if any pause after reinforcement (e.g., gambler who plays the slot machines). A fixed ratio schedule is predictable and produces a high response rate, with a short pause after reinforcement (e.g., salesperson paid on commission). The variable interval schedule is unpredictable and produces a moderate, steady response rate (e.g., restaurant manager who is paid a bonus if the food inspector, who comes at unpredictable times, gives the restaurant a good rating). The fixed interval schedule yields a scallop-shaped response pattern, reflecting a significant pause after reinforcement (e.g., surgery patient who can receive a painkiller after pressing a button, but only up to once per hour). (Figure and caption adapted from @Spielman2020; [https://openstax.org/books/psychology-2e/pages/6-3-operant-conditioning](https://openstax.org/books/psychology-2e/pages/6-3-operant-conditioning); Spielman, R. M., Jenkins, W. J., & Lovett, M. D. (2020). *Psychology*. OpenStax. [https://openstax.org/details/books/psychology-2e](https://openstax.org/details/books/psychology-2e); archived at [https://perma.cc/9DZ8-34X2](https://perma.cc/9DZ8-34X2)).
:::
